Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2015-03-23 draftfinalreport_v2.Rnw
Final edits

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks after measles introductions from abroad. In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identify additional analyses that may help understand risk of infection in New Zealand. Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007 until mid-2014. We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of na\"{\i}ve people at the District Health Board (DHB) and national level. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings are:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item The regression analyses suggest that age is a particularly strong risk factor for measles.  People of European ethnicity make up the majority of cases. However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis. Increased wealth (NZDep 1--3) was a risk factor. Pacific people are at greater risk per capita. However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item Peak overall overseas travel rates, and thus presumably risk from measles importation, is in December. However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed one, including 2013--2014 outbreaks. As analysed from data until June, the last outbreak to end in 2014 had an $R_v$ well above 1 (mean estimate 3.66, range 3.4--3.92). The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92. This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population and that large outbreaks could occur.
\item The cost of 2014 measles cases is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
\item The mean wage loss per measles case is estimated to be approximately \$839.
\item The mean cost per measles case attending hospital is estimated to be \$1,877 per case, and approximately 17\% of measles cases attend hospital
\item The mean public health service cost per case is \$1,765 excluding hospitalisation costs.
\item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170.4 per day.
\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed approximately \$8,214 per vaccine before the costs exceed the benefits.
\item Measles introductions to New Zealand with a median size of 2 cases per year were predicted by simulation following catch up vaccination to ensure $R_v$ is < 1. Thus, typically individual cases and secondary cases will be the expectation. However, the mean of 61 cases was predicted, because peak outbreak sizes of tens, hundreds, and very rarely up to many thousands are predicted following importation, despite $R_v$ being one and the outbreak predicted to die out following additional vaccinations. Thus, increased vaccination beyond the 28\% of the currently 11\% na\'{\i}ve population required may be useful.
\end{itemize}

@@

With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories. Because for measles cases the very young appear to be disproportionately affected (\autoref{fig:ageinyears}), we split the 0--5 age category into additional classes. Following discussion with the Ministry, these were 0--1, 2--4, and 5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}). The 5 year olds were added to the 6--17 year old category. No breakdown of 6--17 year category was performed as constant birth rates for the period could not be assumed.

@@

\begin{sidewaysfigure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{percap_ordert.pdf}
\end{center}
\caption{Per capita measles values from 2007--2014 (\autoref{table:percap}). The x-axis shows NZDep:Age Range:Prioritized ethnicity. Absolute cases are in \autoref{fig:cases_all}}
\label{fig:percap_all}
\end{sidewaysfigure}

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap_all}, \autoref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}). The breakdown of cases and per capita measles values can be see in \autoref{fig:cases_all} and \autoref{fig:percap_all}. Two way interactions are shown in the Appendix (\autoref{sec:app}, \autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).

\begin{table}[hbtp] \scriptsize
<<results=tex,echo=FALSE>>=
colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@
\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
\label{table:percap}
\end{table}

@@

\vspace{5mm} %5mm vertical space
\begin{table}
<<include=FALSE,echo=FALSE,results=tex>>=
  print(xtb,floating = FALSE)
@
\caption{Summary of the regression model results, estimated using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-2014}
\label{table:regression_riskfactors}
\end{table}
\vspace{5mm} %5mm vertical space

@@

The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:cases_all}). Indeed, the majority of cases were in more wealthy European 5--17 year olds (\autoref{fig:cases_all}). However, per capita rates for this large age class were lower than in the very young (0--1 year olds, \autoref{fig:percap_all}). People of Pacific origin are over-represented as measles cases on a per capita basis ($\beta$ = 1.22, standard error (SE) = 0.21, p--value < 0.0001; \autoref{table:regression_riskfactors}), and in particular the 0--1 year old less wealthy (NZDep 8--10) (\autoref{fig:percap_all}. Note however that generally those people in NZDep levels 4--10 are less represented on a per capita basis (NZDep 4--7 $\beta$ = -0.26, SE = 0.09, p--value 0.009; NZDep 8--10 $\beta$ = -0.3, SE = 0.1, p--value == 0.006; \autoref{table:regression_riskfactors}), and in European and Maori ethnicities there is a trend in the opposite way than Pacific peoples, with the more wealthy more likely to be cases (\autoref{fig:percap_all} and \autoref{fig:cases_all}).

There are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 5--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap},\autoref{table:sero}, and \autoref{sub:popim}). Interestingly the regression results suggest risk of measles in the 5--17 year age category was greater for Europeans and Maori (\autoref{table:percap}, \autoref{table:regression_riskfactors}, \autoref{fig:percap_all}).

@@

\begin{table}[htdp]
\begin{center}
\begin{tabular}{lrrrr}
\hline
Age &  Proportion immune\\
\hline
0  & 0\\
1  & 0.89\\
2  & 0.92\\
3	& 0.93\\
4	& 0.93\\
5	& 0.92\\
6-13	& 0.8\\
14-18	& 0.83\\
19-23	& 0.77\\
24-32	& 0.85\\
33-52	& 0.92\\
>52	& 0.99\\
\hline
\end{tabular}
\end{center}
\caption{Vaccination coverage (0--5 years) and serosurvey estimates of immunity ($>5$ years) among different age classes used in these analyses, as provided by the Ministry of Health. Note these values are from vaccination records and serosurveys. Vaccine efficacy for 0--5 year olds was 99\%, 6--13 year olds 96\%, and for \leq 14 equivocal serological results were considered non-immune}
\label{table:sero}
\end{table}%

@@

The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (\autoref{fig:ageandvac}). However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second. A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}). Plotting the cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).

@@

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{vacc_age.pdf}
\end{center}
\caption{Age of vaccination of vaccinated measles cases, 2007--2014. Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (60 months)}
\label{fig:vaccstat}
\end{figure}

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{dob_vacc_unvac.pdf}
\end{center}
\caption{Year of birth and vaccination status of measles cases, 2007--2014}
\label{fig:yearandvac}
\end{figure}

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{dob_vacc_dose.pdf}
\end{center}
\caption{Year of birth and vaccination status of measles cases, 2007--2014}
\label{fig:yearandvacf}
\end{figure}

@@

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{naive_au_prop.pdf}
\end{center}
\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator. Note no values should be greater than 1 (100\% vaccine cover), and the maximum discordance in > 1600\%.}
\label{fig:naive_au_prop}
\end{figure}

@@

\subsection{Measles importation risk methods}
\label{sub:imp_risk}
For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). WHO measles coverage is the percent of children that receive one dose of measles vaccine by their second birthday. Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports. We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from. We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation. We use 0--19 year old data for the immigration figures, as this is the only category available and it was determined this was more appropriate than using all age classes as > 19 year olds are believe to be immune to measles more or less globally. These data were not available for New Zealand travellers, so we use all travel. We present the data from 2013 because this year was the most recent with the most complete WHO measles data, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage. Data for 2013 are incomplete, but substantially better than 2014.

@@

\begin{table}
<<results=tex,echo=FALSE>>=
latex(toprisk2, file="", table.env=FALSE,rowname=NULL)
@
\caption{Measles and travel data from 2013 for the top 10 countries identified as high risk for measles importation. incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of 0--19 year old travellers entering New Zealand from the country}
\label{table:toprisk}
\end{table}

@@

The regression analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at greater risk of measles. In particular, Pacific people are at greater risk per capita, as are the less deprived (NZDep 1--3), and European and Maori 5--17 year old children compared to Asian and Pacific ethnicity children of the same age. Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses. For example, increased numbers of more wealthy Europeans were measles cases, as was true for Maori. However, the alternative situation was clear for Pacific people, especially of the 0--1 year olds (\autoref{fig:percap_all} and \autoref{fig:cases_all}). Results such as this may simply due to the spatial effects due the communities the outbreaks were occuring in, rather than an effect of wealth. A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data. The data suggest that census data and NIR data are recording children living in different DHBs to that in which they are vaccinated, or subsequent to the census or NIR data moving. This leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated. Another issue was how to deal with people of greater age than those recorded in the NIR. Thus, these analyses are possible future directions for this work and a focus of future data collection could be to better understand the link between denominator data and NIR data.

@@

The distribution of na\"{\i}ve among the DHBs also varies (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, \autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).  We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB. However, it is worth bearing in mind that the data here assume that the national NIR vaccination coverage and serosurvey results are appropriate for all DHB.

Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand. However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts. The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}). However, recently Australia was declared free of endemic measles and so the 2013 data must reflect imported measles there. Other interaction terms beyond the product of travel and infection incidence may be more appropriate but this will require substantial further analyses. Other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (\autoref{table:risk12} and \autoref{fig:risk12}). The 2012 data were more complete than the 2013 data, which suffers from data deficiencies, but due to the dynamic nature of measles epidemiology and consultation with the Ministry we present the 2013 data. Lastly, more refined data on the age of travellers may improve the risk assessment, but these are currently unavailable.

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age.
\item European ethnicity comprise the largest number of measles cases.
\item Pacific people are statistically more at risk per capita of measles infection.
\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
\item There is some statistical support for 2--24 year old Pacific people being less at risk than European and Maori of the same age, as are 5--17 year old Asian children.
\item The majority of cases are unvaccinated.
\item The majority of vaccine failures occur in those people which recieved single vaccinations around 1 year old.
\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas.
\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
\item There may be seasonal changes in risk of measles importation, with travel numbers peaking in December.
\end{itemize}

@@

In total, with additional \$20 costs per GP consultation, we estimate that the 2014 measles cases alone cost the New Zealand economy \$1041185.63.

>>>>

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks after measles introductions from abroad. In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identify additional analyses that may help understand risk of infection in New Zealand. Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007 until June 2014. We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of na\"{\i}ve people at the District Health Board (DHB) and national level. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings are:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item The regression analyses suggest that age is a particularly strong risk factor for measles.  Increased wealth (NZDep 1--3) was a risk factor and people of European ethnicity make up the majority of cases. However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis. Pacific people are at greater risk per capita. However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations nationally.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe. Peak overall overseas travel rates, and thus presumably risk from measles importation, is in December. However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed one, including 2013--2014 outbreaks. As analysed from data until June, the last outbreak to end in 2014 had an $R_v$ well above 1 (mean estimate 3.66, range 3.4--3.92). The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92. These analyses suggest improved vaccination is a requisite to prevent measles persisting in the population and that large outbreaks could occur.
\item The cost of 187 measles cases in 2014 is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
\item The mean wage loss per measles case is estimated to be approximately \$839.
\item The mean cost per measles case attending hospital is estimated to be \$1,877 per case, and approximately 17\% of measles cases attend hospital
\item The mean public health service cost per case is \$1,765 excluding hospitalisation costs.
\item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170.4 per day.
\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed approximately \$8,214 per vaccine before the costs exceed the benefits.
\item Following catch up vaccination to ensure $R_v$ is < 1 measles introductions to New Zealand with a median size of 2 cases were predicted by simulation. Thus, typically individual cases and secondary cases will be the expectation. However, the mean outbreak size of 61 cases was predicted, because outbreak sizes of tens, hundreds, and very rarely up to many thousands of cases are predicted among the 1000 simulations following importation, despite $R_v$ being one and the outbreak predicted to die out. Thus, increased vaccination beyond the 28\% of the currently 11\% na\'{\i}ve population required may be useful to prevent these rare but costly events.
\end{itemize}

@@

With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories. Because for measles cases the very young appear to be disproportionately affected (\autoref{fig:ageinyears}), we split the 0--5 age category into additional classes. Following discussion with the Ministry, these were 0--1, 2--4, and 5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}). The 5 year olds were added to the 6--17 year old category to make a 5--17 year old category. No breakdown of 5--17 year category was performed as constant birth rates for the period could not be assumed.

@@

\begin{sidewaysfigure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{percap_ordert.pdf}
\end{center}
\caption{Per capita measles values from 2007--2014 (\autoref{table:percap}). The x-axis shows NZDep:Age Range:Prioritized ethnicity. Numbers of cases are in \autoref{fig:cases_all}}
\label{fig:percap_all}
\end{sidewaysfigure}

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{table:percap}), leading to very large standard error in regression analyses. However, there are also numerous issues with the data for MLA ethnicity category above and beyond the small population sizes, creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}). The breakdown of cases and per capita measles values can be see in \autoref{fig:cases_all} and \autoref{fig:percap_all}. Two way interactions are shown in the Appendix (\autoref{sec:app}, \autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).

\begin{table}[hbtp] \scriptsize
\captionsetup{width=1.2\textwidth}
<<results=tex,echo=FALSE>>=
colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@
\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014. MLA is Middle Eastern/Latin American/African}
\label{table:percap}
\end{table}

@@

\vspace{5mm} %5mm vertical space
\begin{table}
<<include=FALSE,echo=FALSE,results=tex>>=
  print(xtb,floating = FALSE)
@
\caption{Summary of the regression model results using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-June 2014}
\label{table:regression_riskfactors}
\end{table}
\vspace{5mm} %5mm vertical space

@@

The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:cases_all}). Indeed, the majority of cases were in more wealthy European 5--17 year olds (\autoref{fig:cases_all}). However, per capita rates for this large age class were lower than in the very young (0--1 year olds, \autoref{fig:percap_all}). People of Pacific origin are over-represented as measles cases on a per capita basis ($\beta$ = 1.22, standard error (SE) = 0.21, p--value < 0.0001; \autoref{table:regression_riskfactors}), and in particular the 0--1 year old less wealthy (NZDep 8--10) (\autoref{fig:percap_all}). Note however that generally those people in NZDep levels 4--10 are less represented on a per capita basis (NZDep 4--7 $\beta$ = -0.26, SE = 0.09, p--value 0.009; NZDep 8--10 $\beta$ = -0.3, SE = 0.1, p--value == 0.006; \autoref{table:regression_riskfactors}), and in European and Maori ethnicities there is a trend opposite to that in than Pacific people, with the more wealthy more likely to be cases (\autoref{fig:percap_all} and \autoref{fig:cases_all}).

There are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 5--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented than the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap}, \autoref{table:sero}, and \autoref{sub:popim}). Interestingly the regression results suggest risk of measles in the 5--17 year age category was greater for Europeans and Maori (\autoref{table:percap}, \autoref{table:regression_riskfactors}, \autoref{fig:percap_all}).

@@

\begin{table}[htdp]
\begin{center}
\begin{tabular}{lrrrr}
\hline
Age &  Proportion immune\\
\hline
0  & 0\\
1  & 0.89\\
2  & 0.92\\
3	& 0.93\\
4	& 0.93\\
5	& 0.92\\
6-13	& 0.8\\
14-18	& 0.83\\
19-23	& 0.77\\
24-32	& 0.85\\
33-52	& 0.92\\
>52	& 0.99\\
\hline
\end{tabular}
\end{center}
\caption{Vaccination coverage (0--5 years) and serosurvey estimates of immunity ($>5$ years) among different age classes used in these analyses, as provided by the Ministry of Health. Note these values are from vaccination records and serosurveys. Vaccine efficacy for 0--5 year olds was 99\%, 6--13 year olds 96\%, and for $\leq$ 14 year olds equivocal serological results were considered non-immune}
\label{table:sero}
\end{table}%

@@

The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (\autoref{fig:ageandvac}). However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second. A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}). Plotting the unvaccinated cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).

@@

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{vacc_age.pdf}
\end{center}
\caption{Age of vaccination of vaccinated measles cases, 2007- June 2014. Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (60 months)}
\label{fig:vaccstat}
\end{figure}

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{dob_vacc_unvac.pdf}
\end{center}
\caption{Year of birth of unvaccinated of measles cases, 2007- June 2014}
\label{fig:yearandvac}
\end{figure}

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{dob_vacc_dose.pdf}
\end{center}
\caption{Year of birth of unvaccinated of measles cases, 2007- June 2014. The year of birth for those vaccinated with measles, mumps, and rubella (MMR) dose 1 is shown in the upper panel, MMR dose 2 the lower panel.}
\label{fig:yearandvacf}
\end{figure}

@@

\begin{figure}
\begin{center}
     \includegraphics[width=0.9\textwidth]{naive_au_prop.pdf}
\end{center}
\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator. Note no values should be greater than 1 (100\% vaccine cover), and the maximum discordance is > 1600\%.}
\label{fig:naive_au_prop}
\end{figure}

@@

\subsection{Measles importation risk methods}
\label{sub:imp_risk}
For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). WHO measles coverage is the percent of children that receive one dose of measles vaccine by their second birthday. Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports. We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from. We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation. We use 0--19 year old data for the immigration figures, as no further details were available and it was determined this was more appropriate than using all age classes because > 19 year olds are believe to be immune to measles more or less globally. These age category data were not available for New Zealand travellers, so we use all data on all travellers. We present the data from 2013 because this year was the most recent with the relatively complete WHO measles data, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage. Data for 2013 are less complete than 2012, but substantially better than 2014.

@@

\begin{table}
<<results=tex,echo=FALSE>>=
latex(toprisk2, file="", table.env=FALSE,rowname=NULL)
@
\caption{Measles and travel data from 2013 for the top 10 countries identified as high risk for measles importation. incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of New Zealander and 0--19 year old foreign national travellers entering New Zealand from the country}
\label{table:toprisk}
\end{table}

@@

The regression analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at greater risk of measles. In particular, Pacific people are at greater risk per capita, as are the less deprived (NZDep 1--3), and European and Maori 5--17 year old children compared to Asian and Pacific ethnicity children of the same age. Furthermore, Pacific origin people of all categories from 2--24 years old were less likely on a per capita basis to be cases than Europeans or Maori. Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses. For example, increased numbers of more wealthy Europeans were measles cases, as was true for Maori. However, the alternative situation was clear for Pacific people, especially of the 0--1 year olds (\autoref{fig:percap_all} and \autoref{fig:cases_all}). Results such as this may simply be due to the spatial effects in the communities the outbreaks were occuring in, rather than an effect of wealth. A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data. The data suggest that census data and NIR data are recording children living in different DHBs to that in which they are vaccinated, or subsequent to the census or NIR data moving. This leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated. Another issue was how to deal with people of greater age than those recorded in the NIR. Thus, these analyses are possible future directions for this work and a focus of future data collection could be to better understand the link between denominator data and NIR data.

@@

The distribution of na\"{\i}ve among the DHBs also varies, though the proportion of na\"{\i}ve overall across DHBs is relatively similar (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, \autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).  We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB. However, it is worth bearing in mind that the data here assume that the national NIR vaccination coverage and serosurvey results are appropriate for all DHB.

Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand. However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts. The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}). Recently Australia was declared free of endemic measles and so the 2013 data must reflect imported measles there. Other interaction terms beyond the product of travel and infection incidence may be more appropriate but this will require substantial further analyses. Other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (\autoref{table:risk12} and \autoref{fig:risk12}). The WHO 2012 data were more complete than the WHO 2013 data, which suffers from data deficiencies, but due to the dynamic nature of measles epidemiology and consultation with the Ministry we present the WHO 2013 data. Lastly, more refined data on the age of travellers may improve the risk assessment, but these are currently unavailable.

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age.
\item European ethnicity comprise the largest number of measles cases.
\item Pacific people are statistically more at risk per capita of measles infection.
\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
\item There is statistical support for 2--24 year old Pacific people being less at risk than European and Maori of the same age, as are 5--17 year old Asian children.
\item The majority of cases are unvaccinated.
\item The majority of vaccine failures occur in those people which recieved single vaccinations around 1 year old.
\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas, though the \%  na\"{\i}ve is similar across DHBs.
\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
\item There may be seasonal changes in risk of measles importation, with travel numbers peaking in December.
\end{itemize}

@@

In total, with additional \$20 costs per GP consultation, we estimate that the 187 measles cases in 2014 cost the New Zealand economy \$1041185.63.
